Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-22 examples_C.tex
major revision of explanation up through extended invariant, with concrete plans up through (not including) walkpgdir discussion

  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{entry}\},m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$

@@

@@

    (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$ (*\label{line:after_concluding_qfrac1C}*)

@@

    /*Now we know that entry is initialized, so we satisfy the condition to access children list */

@@

    $\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (pfn\_set(entry\_val nextpaddr)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$ (*\label{line:alloc_path_endC}*)

@@

\subsubsection{From L$n+1$ Entries to L$n$ Tables}
We will discuss access to the level 4 table later (Section \ref{wlkpgdirC}). However, for subsequent levels, the base address of level $n$ must be
fetched from the appropriate entry in the table of level $n+1$.
This is the role of \lstinline|pte_get_next_table| (originally Figure \ref{fig:pagetablescode}, with proof details in Figure \ref{fig:calltopteinitializeC}).
It is passed the virtual address of the page table entry in level $n+1$, and should return the \emph{virtual} 
address of the \emph{base} of the level $n$ table
indicated by that entry.
If the entry is empty (i.e., this is a sparse part of the page table representation),
the code also allocates a page for the level $n$ table, installs it in the level $n+1$ entry, and establishes appropriate invariants.
Figure \ref{fig:calltopteinitializeC} presents the function with proof annotations that we will explain shortly, but we first explain the functionality.
\lstinline|pte_get_next_table| accepts a \emph{virtual} address \lstinline|entry| which points to a level-$n+1$
table entry. On Line \ref{line:read_entry_contentsC}, the code checks the present bit of the entry.
If the bit is unset, there is no level-$n$ table, so one must be allocated via \lstinline|pte_initialize| (explained shortly\footnote{
  Briefly, it allocates a fresh physical page, and initializes the memory pointed to by \lstinline|entry| with that physical address.
}) and marked present.
By Line \ref{line:finalpieceS} the entry is known to be valid and contain the physical address of
the base of a level $n$ table. That address is then extracted (Line \ref{line:extract_pfn}),
converted to a virtual address (Line \ref{line:p2vC}), and returned to the caller.

We can now discuss \lstinline|pte_get_next_table|'s proof of correctness, which is surprisingly subtle for two
local reasons, and a third related to interaction with the caller, none of which have been addressed by prior work.
First, the update to the memory at (virtual) address \lstinline|entry| depends on subtle ownership invariants:
if the entry is present then its fractional ownership is shared with a large number of \lstinline|L4_L1_PointsTo| assertions
from the address space invariant (Figure \ref{fig:peraspaceinvariant}),
but if the entry is absent the proof requires full ownership to update it. We resolve this by extending the address space invariant
to make the owned fraction of the entry's memory \emph{dependent on its own contents}.
Second, the conversion of physical addresses into a corresponding virtual address that can be used to modify the specific
physical location relies on subtle, never-before-formalized kernel invariants.
And third, these two factors percolate to the precondition (for conditional fractional ownership) and postcondition
(for physical-to-virtual conversions), as the caller essentially passes output from one call to \lstinline|pte_get_next_table|
as input to a subsequent call to traverse multiple levels.
Since the key to solving these challenges is to extend the address space invariant, we
first discuss that invariant and the kernel designs it supports, before returning to the subtle details of
verifying lines \ref{line:install_new_entryC} and \ref{line:p2vC}.

\subsubsection{An Extended Address Space Invariant: Identity Mappings and Conditional Ownership Page Table Ownership}
\label{subsec:identitymappingsC}

\ldots
\hrule\hrule

Kernels need to convert between physical and virtual addresses, in both directions.
Traversing the page tables in software is the simplest way to convert a virtual address to a physical address; this is the context we are working up to.
However, implementing this virtual-to-physical (V2P) translation in this way ironically requires physical-to-virtual (P2V) translation,
because the addresses stored in page table entries are physical, but memory accesses issued by the OS code use virtual addresses.
% There is no universal way to convert physical addresses to virtual --- doing so relies on the kernel maintaining careful invariants or
% additional data structures to enable P2V translation.
\looseness=-1

Because VMM operations are performance-critical for many workloads, most kernels
maintain invariants that enable very fast P2V conversions (rather than adding another data structure).
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table 
% lives at a virtual address whose value 
is \emph{a constant offset from the physical address} --- a practice sometimes referred to as \emph{identity mapping} 
(even though the physical-to-virtual translation
is typically not literally the identity function, but adding a nonzero constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.
On newer platforms like RISC-V, this sometimes truly is an identity mapping ---
x86-64 machines are forced into offsets by backward compatibility with bootloaders that cannot access the full memory space of the
machine.
}

For this reason, we extend the per-address-space invariant as in Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC}, to also track which
addresses we can perform a P2V conversion on by adding a constant offset.
$\Xi$ is another ghost map, from physical addresses to the level of the page table they represent (1--4).
\emph{Only} physical addresses in $\Xi$ can undergo P2V conversion. 
Section \ref{sec:p2vC} describes the actual conversion,
but we describe the invariant here because adding new level 3/2/1 tables must maintain the invariant.

\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast \mathsf{GhostMap}(\mathsf{id},\Xi)\ast\\
  \left(\bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})}\right)\ast \\
  \bigast{(\paddr,\mathsf{level}) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{KERNBASE} \; \textsf{level} > 1\urcorner \ast  \underbrace{\fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} }_\text{Ghost translation} \ast \underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location}} \ast\\
   \qquad\underbrace{ \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}_\text{Entry validity}\ast 
    \underbrace{\left(\ulcorner\textsf{present\_L}(\vale,\mathsf{level})\urcorner \wand \forall_{\textsf{i}\in\textsf{0..511}} \ldotp \ghostmaptoken{\textsf{id}}{((\mathsf{entry\_page}\;\vale) + \textsf{i * 8})}{\textsf{level-1}}\right)}_{\text{Indexing into next level of tables}} \\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   % \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}
   \ptableabswalk{\delta,\ptablestore} \ast \pvmapping{\delta,\Xi}\\
  \textsf{present\_L}(\vale,\mathsf{level})\stackrel{\triangle}{=} \mathsf{entry\_present}(\vale)\land \mathsf{level} > 0

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariantC} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extensionC}
\vspace{-1em}
\end{figure*}

For each $\paddr\mapsto \textsf{v} \in\Xi$, the invariant tracks virtual points-to justifying that virtual address $\paddr+\textsf{KERNBASE}$ maps to physical address $\paddr$
(the ``Ghost translation'' in Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC});
fractional ownership of the physical memory for that page table entry;
and for valid entries (with the present bit set) above L1, ghost map tokens for every entry in the table pointed to by the entry, which can be used
to repeat the process one level down. 
% (L1 entries point to data pages, whose physical memory ownership resides in some virtual points-to).
The assertion on Line \ref{line:conditional_childrenC} of Figure \ref{fig:calltopteinitializeC} comes from the invariant one level up; 
if the valid bit is set,
the code can return those child tokens without the conditional guard.
\looseness=-1

The fractional ownership of the entry's physical memory is subtle. Recall that $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ retains some physical
ownership of each page table entry that is traversed (proportional to how many virtual addresses share the entry).
So in general the invariant cannot keep full permission to the memory in this part of the invariant, or it would overlap the page table walk for virtual points-to
assertions. But in the case where the entry is invalid, we may need to write to it (e.g., to install a reference to a next-level table, as we do in Figure \ref{fig:calltopteinitializeC}),
which requires full permission. Fortunately, the entry can only be in use if its valid bit is set; if the valid bit is not set, we know
that no virtual points-to entry in $\delta$/$\theta$ holds any partial ownership.
Thus we use the invariant portion annotated as ``Entry validity'' in Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC} to capture this:
if the entry is invalid the invariant holds full ownership of the entry, so it can be updated; while if the entry is valid,
the invariant owns only a constant nonzero fragment sufficient to read the entry, but not modify it (which would invalidate some virtual points-to assertions):
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
Thus the fractional ownership of the physical location is enough for Line \ref{line:read_entry_contentsC} in Figure \ref{fig:calltopteinitializeC} to access the entry, though in \lstinline|get_next_table|
the caller has pulled that piece of information out of the invariant and passed it for the entry at hand.
This removal appears explicitly in assertions,
as the argument to the invariant is $\Xi\setminus\{\mathsf{entry}\}$ (indexing by the set $\Xi$ allows us to borrow the physical resources
for a specific page table entry out of the invariant, and later put them back).
Line \ref{line:check_entry_present_jumpC}'s conditional then determines in the fall-through case that the bit is not set, which
together with other facts entails $\textsf{qfrac} = 1$ at Line \ref{line:after_concluding_qfrac1C},
and permits storing a new entry (in the elided code around Line \ref{line:install_new_entryC}).
\looseness=-1

This seemingly simple piece of code has a highly nontrivial correctness argument, which depends critically on detailed invariants on how access to page table
entries is shared between parts of the kernel. No prior work has engaged with this problem.

@@

 \subsubsection{Installing a New Table}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
\ifPLDI
Line \ref{line:mask_presentC} % in condensed version, all on same line
\else
Lines \ref{line:mask_present}--\ref{line:check_entry_present} 
\fi
in Figure \ref{fig:calltopteinitializeC}.
Accessing the presence bit and checking the value allows us to exploit the condition (*) that was just discussed when verifying the allocation
path (i.e., when the entry is invalid and Lines \ref{line:alloc_path_startC}--\ref{line:alloc_path_endC} in Figure \ref{fig:calltopteinitializeC}
must allocate the next level of tables).
This operation is subtle. To reiterate: the operation requires that the relevant table entry is readable, but the exact portion of ownership
returned must be determined by inspecting the valid bit of the value in memory --- so full ownership is returned only for unused entries.
When the bit is not set, that entails full ownership of the entry's memory ($\textsf{qfrac} = 1$) and justifies writing to that memory.
Otherwise, the code jumps past the end of this listing, to the following code at the top of Figure \ref{fig:p2v} (which is also the
continuation of this code).

@@

\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
/*
@param l4: the root address of the page-table tree
@param va: virtual address to be translated and mapped
*/

@@

@@

\begin{figure}\footnotesize
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{l3}+\textsf{L3Offset(va)} = \textsf{l3\_entry} \land \textsf{table\_root(l4\_entry}_{\textsf{pfn}})  = \textsf{l3} \\ \land \textsf{l2}+\textsf{L2Offset(va)} = \textsf{l2\_entry} \land \textsf{table\_root(l3\_entry}_{\textsf{pfn}})  = \textsf{l2} \\ \land \textsf{l1}+\textsf{L1Offset(va)} = \textsf{l1\_entry} \land \textsf{table\_root(l2\_entry}_{\textsf{pfn}})  = \textsf{l1} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{l4\_entry},\textsf{l3\_entry},\textsf{l2\_entry},\textsf{l1\_entry}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  
      \ghostmaptoken{\textsf{id}}{(\mathsf{l4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{l3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{l2\_entry})}{\textsf{2}}  \ast \\  \textsf{l4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4\_entry} \; \textsf{l4e\_val}  \ast \\  \textsf{l3\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3\_entry}\;\textsf{l3e\_val}  \ast \\  \textsf{l2\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2\_entry} \; \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } 
    \\ \ast \left(\begin{array}{l}  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry}\;\textsf{l1e\_val} \ast \\ \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l1e\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{l1e\_get\_next\_table} in Figure \ref{fig:calltopteinitializeC}}
\label{fig:rwalkC}
\vspace{-1em}
\end{figure}
Implementing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdirC}. 
The key part of the specification and proof for a page table walk is accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpteC} and \ref{line:ex_l3_vpteC} in Figure \ref{walkpgdirC} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

>>>>

  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{entry}\},m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$(*\label{line:precondition_entry_out}*)

@@

  $\specline{\ulcorner\textsf{entry\_present(entry\_val)}\urcorner \wand \forall_{i\in\textsf{0..511}} \; \ghostmaptoken{\textsf{id}}{\textsf{table\_root(entry\_val.pfn) + i * 8}}{\textsf{v-1}} }_{\rtv} $(*\label{line:precondition_conditional}*) 

@@

    (*\label{line:refined_fractional_ownership}*)(*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$ (*\label{line:after_concluding_qfrac1C}*)

@@

    /*Now we know that entry is initialized, so we satisfy the condition to access children list */(*\label{line:now_we_know}*)

@@

    $\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (pfn\_set(entry\_val nextpaddr)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$(*\label{line:available_child_tokens}*) (*\label{line:alloc_path_endC}*)

@@

\subsubsection{From L$n+1$ Entries to L$n$ Tables}
We will discuss access to the level 4 table later (Section \ref{wlkpgdirC}). However, for subsequent levels, the base address of level $n$ must be
fetched from the appropriate entry in the table of level $n+1$.
This is the role of \lstinline|pte_get_next_table| (originally Figure \ref{fig:pagetablescode}, with proof details in Figure \ref{fig:calltopteinitializeC}).
It is passed the virtual address of the page table entry in level $n+1$, and should return the \emph{virtual} 
address of the \emph{base} of the level $n$ table
indicated by that entry.
If the entry is empty (i.e., this is a sparse part of the page table representation),
the code also allocates a page for the level $n$ table, installs it in the level $n+1$ entry, and establishes appropriate invariants.
Figure \ref{fig:calltopteinitializeC} presents the function with proof annotations that we will explain shortly, but we first explain the functionality.
\lstinline|pte_get_next_table| accepts a \emph{virtual} address \lstinline|entry| which points to a level-$n+1$
table entry. On Line \ref{line:read_entry_contentsC}, the code checks the present bit of the entry.
If the bit is unset, there is no level-$n$ table, so one must be allocated via \lstinline|pte_initialize| (explained shortly\footnote{
  Briefly, it allocates a fresh physical page, and initializes the memory pointed to by \lstinline|entry| with that physical address.
}) and marked present.
By Line \ref{line:finalpieceS} the entry is known to be valid and contain the physical address of
the base of a level $n$ table. That address is then extracted (Line \ref{line:extract_pfn}),
converted to a virtual address (Line \ref{line:p2vC}), and returned to the caller.
We can now discuss \lstinline|pte_get_next_table|'s proof of correctness.
While at first glance this code does not sound terribly complex, beyond needing some care to distinguish physical and virtual addresses,
it has a highly nontrivial correctness argument, which depends critically on detailed invariants on how access to page table
entries is shared between parts of the kernel. No prior work has engaged with this problem.

\subsubsection{Address Space Invariant: Identity Mappings and Conditional Ownership Page Table Ownership}
\label{subsec:identitymappingsC}
Assembly-level verification of compiler output from Figure \ref{fig:calltopteinitializeC} is verbose but doable
Section \ref{sec:logic}'s logic, but only after resolving two critical challenges.\todo{Ismail, not sure if this ``verbose but doable''
 is a good framing, we should discuss
}
Two key challenges stand out and end up affecting both the pre- and post-conditions, neither of which has been addressed by prior work.
First, the update to the memory at (virtual) address \lstinline|entry| depends on subtle ownership invariants:
if the entry is present then its fractional ownership is shared with a large number of \lstinline|L4_L1_PointsTo| assertions
from the address space invariant (Figure \ref{fig:peraspaceinvariant}),
but if the entry is absent the proof requires full ownership to update it. We resolve this by extending the address space invariant
to make the owned fraction of the entry's memory \emph{dependent on its own contents}.
Second, the conversion of physical addresses into a corresponding virtual address that can be used to modify the specific
physical location relies on subtle, never-before-formalized kernel invariants.
% \looseness=-1
%
% These two factors percolate to the precondition (for conditional fractional ownership) and postcondition
% (for physical-to-virtual conversions), as the caller essentially passes output from one call to \lstinline|pte_get_next_table|
% as input to a subsequent call to traverse multiple levels.
Since the key to solving these challenges is to extend the address space invariant, we
first discuss that invariant and the kernel designs it supports, before returning to the subtle details of
verifying lines \ref{line:install_new_entryC} and \ref{line:p2vC}.

\paragraph{Physical-to-Virtual Mappings}
Kernels need to convert between physical and virtual addresses, in both directions.
Traversing the page tables in software is the simplest way to convert a virtual address to a physical address;
this is the context we are working up to.
However, implementing this virtual-to-physical (V2P) translation in software ironically requires physical-to-virtual (P2V) translation,
because the addresses stored in page table entries are physical, but memory accesses issued by the OS code use virtual addresses.
Because VMM operations are performance-critical for many workloads, most kernels
maintain invariants that enable very fast P2V conversions (rather than adding another data structure).
Specifically, many kernels maintain an invariant on their page tables that the virtual address of any page used for a page table 
% lives at a virtual address whose value 
is \emph{a constant offset from the physical address} --- a practice sometimes referred to as \emph{identity mapping} 
(even though the physical-to-virtual translation
is typically not literally the identity function, but adding a nonzero constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.
On newer platforms like RISC-V, this sometimes truly is an identity mapping ---
x86-64 machines are forced into offsets by backward compatibility with bootloaders that cannot access the full memory space of the
machine.
}
Thus \lstinline|P2V| on line \ref{line:p2vC} of Figure \ref{fig:calltopteinitializeC} is a macro for adding the fixed constant \lstinline|KERNBASE|.

Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC} extends the per-address-space invariant  to also track which
addresses we can perform a P2V conversion on by adding a constant offset (i.e., the set of physical addresses which participate in page tables).
$\Xi$ is another ghost map, from physical addresses to the level of the page table they represent (1--4).
\emph{Only} physical addresses in $\Xi$ can undergo P2V conversion. 
Section \ref{sec:p2vC} describes the verification of an actual conversion,
but this invariant must be \emph{established} when adding a new page table level (notably on Line \ref{line:call_to_pte_initializeC},
hence the comment of Line \ref{line:now_we_know}).

\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast \mathsf{GhostMap}(\mathsf{id},\Xi)\ast\\
  \left(\bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})}\right)\ast \\
  \bigast{(\paddr,\mathsf{level}) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{KERNBASE} \; \textsf{level} > 1\urcorner \ast  \underbrace{\fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} }_\text{\textcircled{1} Ghost translation} \ast \underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{\textcircled{2} Physical location}} \ast\\
   \qquad\underbrace{ \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner}_\text{\textcircled{3} Entry validity}\ast 
    \underbrace{\left(\ulcorner\textsf{present\_L}(\vale,\mathsf{level})\urcorner \wand \forall_{\textsf{i}\in\textsf{0..511}} \ldotp \ghostmaptoken{\textsf{id}}{((\mathsf{entry\_page}\;\vale) + \textsf{i * 8})}{\textsf{level-1}}\right)}_{\text{\textcircled{4} Indexing into next level of tables}} \\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   % \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}
   \ptableabswalk{\delta,\ptablestore} \ast \pvmapping{\delta,\Xi}\\
  \textsf{present\_L}(\vale,\mathsf{level})\stackrel{\triangle}{=} \mathsf{entry\_present}(\vale)\land \mathsf{level} > 0

\end{array}
\]
\todo[inline]{Ismail, what's APVMappings?}
\vspace{-1em}
\caption{Address space invariant of Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings. }
  \label{fig:peraspaceinvariant_with_p2v_extensionC}
\vspace{-1em}
\end{figure*}

For each $\paddr\mapsto \textsf{v} \in\Xi$, the invariant contains a virtual points-to justifying that virtual address
$\paddr+\textsf{KERNBASE}$ maps to physical address $\paddr$
(\textcircled{1} in Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC});
fractional ownership of the physical memory for that page table entry (\textcircled{2}, which together with \textcircled{1} is equivalent
to a virtual points-to);
and for valid entries (with the present bit set) above L1, ghost map tokens for $\Xi$ for every entry in the table pointed to by the entry, which can be used
to repeat the process one level down (\textcircled{4}). 
% (L1 entries point to data pages, whose physical memory ownership resides in some virtual points-to).
\textcircled{4} becomes part of the precondition to \lstinline|pte_get_next_table|:
Line \ref{line:precondition_conditional}) says that if the entry is valid (points to a next-level table)
then there are tokens for accessing $\Xi$ for every entry in the next-level table.
By Line \ref{line:finalpieceS} the entry is guaranteed to be valid so all tokens for converting the next table level's physical addresses to virtual
are available (Line \ref{line:available_child_tokens}).
\looseness=-1

\paragraph{Self-Conditional Fractional Ownership}
The fractional ownership of the entry's physical memory is subtle.
As noted above, a \emph{valid} entry must coexist with the fractional ownership from
$\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ and therefore have less than full ownership,
but in the case where the entry is \emph{invalid}, Line \ref{line:call_to_pte_initializeC} must have full permissions in order
to populate the entry (i.e., to install a reference to a next-level table).
Fortunately, the entry can only be in use if its valid bit is set; if the valid bit is not set, we know
that no virtual points-to entry in $\delta$ or $\theta$ holds any partial ownership.
But determining this requires reading the very memory whose ownership is being determined.
We use the invariant portion annotated as ``Entry validity'' (\textcircled{3}) in Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC} to capture this:
if the entry is invalid the invariant holds full ownership of the entry, so it can be updated;
while if the entry is valid, the invariant owns only a constant nonzero fraction sufficient to read but not modify the entry.
% \begin{equation*}
%  \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
% \end{equation*}
Since the fractional ownership is always non-zero, Line \ref{line:read_entry_contentsC} in Figure \ref{fig:calltopteinitializeC} can read the entry,
and if the entry is dynamically found to be invalid, the invariant is refined (Line \ref{line:refined_fractional_ownership}) for full
ownership, allowing updates.

Note that the caller is responsible for providing this conditional ownership, having pulled it out of the invariant earlier.
This is why the precondition (Line \ref{line:preconditoin_entry_out}) explicitly excludes the entry's physical address from the invariant ($\Xi\setminus\{\mathsf{entry}\}$) ---
its relevant assertions have already been borrowed by the caller.
\looseness=-1

@@

 \subsubsection{Installing a New Table}
\todo[inline]{C edit frontier. Should see if we can condense this discussion, since the C is only 2 lines and mostly follows from assumptions
about kalloc. Maybe this should get inlined into the Self-Conditional Fractional Ownership above, and likewise the P2V subsubsection below into the Physical-to-Virtual Mappings
bit above. Yeah, they should. I think the expository order still makes sense even if it's not program order; maybe explain we treat the
success case first and the backup case second? This would then still permit the discussion of pte initialize to be near the proof of its call}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
\ifPLDI
Line \ref{line:mask_presentC} % in condensed version, all on same line
\else
Lines \ref{line:mask_present}--\ref{line:check_entry_present} 
\fi
in Figure \ref{fig:calltopteinitializeC}.
Accessing the presence bit and checking the value allows us to exploit the condition (*) that was just discussed when verifying the allocation
path (i.e., when the entry is invalid and Lines \ref{line:alloc_path_startC}--\ref{line:alloc_path_endC} in Figure \ref{fig:calltopteinitializeC}
must allocate the next level of tables).
This operation is subtle. To reiterate: the operation requires that the relevant table entry is readable, but the exact portion of ownership
returned must be determined by inspecting the valid bit of the value in memory --- so full ownership is returned only for unused entries.
When the bit is not set, that entails full ownership of the entry's memory ($\textsf{qfrac} = 1$) and justifies writing to that memory.
Otherwise, the code jumps past the end of this listing, to the following code at the top of Figure \ref{fig:p2v} (which is also the
continuation of this code).

@@

/* @param l4: the root address of the page-table tree
   @param va: virtual address to be translated and mapped */

@@

\todo[inline]{clarify virt/phys of input root and the output}

@@

\begin{figure}\footnotesize
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{l3}+\textsf{L3Offset(va)} = \textsf{l3\_entry} \land \textsf{table\_root(l4\_entry}_{\textsf{pfn}})  = \textsf{l3} \\ \land \textsf{l2}+\textsf{L2Offset(va)} = \textsf{l2\_entry} \land \textsf{table\_root(l3\_entry}_{\textsf{pfn}})  = \textsf{l2} \\ \land \textsf{l1}+\textsf{L1Offset(va)} = \textsf{l1\_entry} \land \textsf{table\_root(l2\_entry}_{\textsf{pfn}})  = \textsf{l1} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{l4\_entry},\textsf{l3\_entry},\textsf{l2\_entry},\textsf{l1\_entry}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  
      \ghostmaptoken{\textsf{id}}{(\mathsf{l4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{l3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{l2\_entry})}{\textsf{2}}  \ast \\  \textsf{l4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4\_entry} \; \textsf{l4e\_val}  \ast \\  \textsf{l3\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3\_entry}\;\textsf{l3e\_val}  \ast \\  \textsf{l2\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2\_entry} \; \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } 
    \\ \ast \left(\begin{array}{l}  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry}\;\textsf{l1e\_val} \ast \\ \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l1e\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\todo[inline]{Colin: This is confusing: in the code \lstinline|l2_entry| is an actual entry word, but in these invariants they seem to be paddrs.}
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{l1e\_get\_next\_table} in Figure \ref{fig:calltopteinitializeC}}
\label{fig:rwalkC}
\vspace{-1em}
\end{figure}
Implementing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdirC}. 
The key part of the specification and proof for a page table walk is accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpteC} and \ref{line:ex_l3_vpteC} in Figure \ref{walkpgdirC} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.
